---
title: "data_manipulation"
output: github_document
date: "2025-09-18"
---

```{r}
library(tidyverse)
```


Import datasets:

```{r}
litters_df =
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

litters_df=
  janitor::clean_names(litters_df)

pups_df = 
    read_csv("data/FAS_pups.csv", skip = 3,na = c("NA", ".", ""))

pups_df=
    janitor::clean_names(pups_df)
```

## Select

`select`  selects variables from a dataframe

Select Specific Variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

Select a range of variables 
```{r}
select(litters_df, group:gd_of_birth)
```

Select by Removal
```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

Select by prefix (`starts_with` is what he uses most often, but there also `ends_with` or `contains`)
```{r}
select(litters_df, group, starts_with("gd"))
```

Rearrange by selecting.

What comes first in the command is what is going to come first in the output. 
```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())
#^second command outputs litternumber first and then outputs everything first

relocate(litters_df, litter_number)
#this does the exact same thing as command two in this section
```

Rename by selecting
```{r}
select(litters_df, GROUP = group, everything())
#GROUP is the new name for group
rename(litters_df, GROUP = group)
#Also does the same thing
```

What if I only select one variable...?
```{r}
select(litters_df, group)
#this is not a vector. It is still a dataframe. Unlike what pull() produces...

pull(litters_df, group)
#he hardly ever ues pull (pulling something out of a dataframe)
#pull produces a vector

```

Theoretically you can do something like `litters_df$group` which pulls the variable out of the dataframe. But this is a clunky, BAD FORM OF CODING

### Learning Assessment 

```{r}
select(pups_df, litter_number:pd_ears)

#Alternatively you can do:
select(pups_df, litter_number, sex, pd_ears)
```




