---
title: "Data Import"
output: github_document
---

```{r}
library(tidyverse)
#readr is part of tidyverse
```


Let's import a dataset. 

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv")

#dont use read.csv it is not as good as read_csv
#by default read_csv assumes first row is header
```

Always look at your data

```{r}
names(litters_df)
```

Accessing variable names that have spaces in it is very hard. 
So update the names in `litter_df`.

```{r}
litters_df=
  janitor::clean_names(litters_df)

#this makes everything snakecase. snake case makes it easier to use variable names in code
```


Look at the data. Most frequent thing jeff does is print the data and view the data set (from console) 

```{r}
litters_df
#<dbl> means that it is a double numeric value
#you can use (view litters_df in console to get a pop up of dataset)
```


Sometimes skimming data is neat?
```{r}
skimr::skim(litters_df)
```


Check your data in your original CSV file to make sure you imported it correctly. 
When we go to CSV, we see that  `gd0_weight` and `gd18_weight` have periods in some of the blank cells. That is why column is read as a `<char>` not a number. 

So we need to fix the missingness. 

```{r}
litters_df=
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))

#now when we run litters_df we will see that `gd0_weight` and `gd18_weight` is now <dbl> not <char>
```


#Import Pups Data
Now import FAS_pups.csv dataset 
  Why it might be helpful to skip the first ten lines: 
  
```{r}
pups_df=
  read_csv("data/FAS_pups.csv", 
           skip = 3,
           na = c("NA", ".", "")
           )

# skip allows you to skip rows 

pups_df=
  janitor::clean_names(pups_df)
```


